Systemix MAG-SMAG Profiling

### 2024-08-22 R/paper.Rmd
improved figures & added abundances

suppressMessages(library(ape)) #tree basics
suppressMessages(library(treeio)) #join trees
suppressMessages(library(ggtree)) #tree plotting
suppressMessages(library(ggplot2)) #plots
suppressMessages(library(ggtreeExtra)) #geom_fruit
suppressMessages(library(ggnewscale)) #scales in ggplot2
suppressMessages(library(ggstar)) #geom_star for ggplot2
suppressMessages(library(tidyverse)) #strings, dataframes processing
suppressMessages(library(magrittr)) #pipes %>%
suppressMessages(library(tidytree)) #trees as dataframes
suppressMessages(library(RColorBrewer)) #color palettes
suppressWarnings(library(colorspace)) #colors manipulation
suppressWarnings(library(patchwork)) #side-by-side plotting
suppressWarnings(library(openxlsx)) #read excel
suppressWarnings(library(ComplexHeatmap)) #heatmap
suppressWarnings(library(PCAtools)) #pca
suppressWarnings(library(ggbiplot)) #pca plots
suppressWarnings(library(agricolae)) #fisher's test
suppressWarnings(library(ggpubr)) #stats & plots
suppressWarnings(library(dndscv)) #dN/dS of snps
suppressWarnings(library(clusterProfiler)) #gene enrichment
suppressWarnings(library(circlize)) #circular viz
```

## Prepare and merge MAG-SMAG metadata

@@

### In-depth biomining MAGs vs SMAG comparison

@@

### QC

@@

### Phylogenetic tree

@@

@@

```{r fig1}
# Merge plots using patchwork
fig_1 <- wrap_plots(plot_mags_qc, plot_depth_sgbs_biom, plot_tree_s15_md,
                    design = "\nACCC\nBCCC", heights = c(0.85, 1)) +
  plot_annotation(tag_levels = list(c("a", "b", "c"))) &
  theme(plot.tag = element_text(size = 18, face = "bold"))
ggsave("article_profiling/fig_1.pdf", fig_1, height = 8, width = 15)
ggsave("article_profiling/fig_1.png", fig_1, height = 8, width = 15)

fig_2 <- wrap_plots(plot_depth_sgbs_both, plot_depth_spec_eco,
                    plot_depth_kggbs_biom, design = "\nAAABBCC\n") +
  plot_annotation(tag_levels = list(c("a", "", "b", "c"))) &
  theme(plot.tag = element_text(size = 18, face = "bold"))
ggsave("article_profiling/fig_2.pdf", fig_2, height = 8, width = 16)
ggsave("article_profiling/fig_2.png", fig_2, height = 8, width = 16)

@@

## Section 2: Functional Annotation

### Pre-process matrix of KEGG KOs from KOfamScan

@@

### Heatmap

@@

### PCA (**discarded**)

@@

### Bubble and boxplots

@@

## Section 3: Evolutionary analysis

### dN/dS

@@

# Violin plot for NS genes with global mutations
snp_plot_nsgenes <- ggviolin(snp_nsgenes, x = "pathway", y = "wmis_cv", color = "pathway",
                             shape = "pathway", xlab = "Pathway", ylab = "dN/dS",
                             order = snp_sort_ns$pathway, palette = snp_sort_ns$color,
                             size = 0.75, add = "jitter", legend = "none",
                             add.params = list(size = 2, alpha = 0.75)) +
  scale_y_continuous(limits = c(-0.2, 1.5)) +
  geom_hline(yintercept = 1, linetype = 1, linewidth = 0.4) +
  geom_hline(yintercept = mean(snp_nsgenes$wmis_cv), linetype = 2, linewidth = 0.4) +
  stat_compare_means(method = "anova", label.y = 1.5, size = 6) +
  stat_compare_means(comparisons = list(c("Sulfur", "Copper"), c("Sulfur", "Iron"),
                                        c("Copper", "Iron")), label = "p.format",
                     method = "t.test", size = 5) +
  theme(axis.text = element_text(size = 16),
        axis.title = element_text(size = 20))
ggsave("article_profiling/5-SNPs/dndscv/results/plot/plot_dnds_nsgenes.pdf", snp_plot_nsgenes, height = 10, width = 7)

@@

# Lollipop plot of grouped NS gene products
snp_plot_nsgensub <- ggdotchart(snp_nsgenes_sub, x = "product", y = "dnds", color = "pathway",
                                xlab = "Product of genes under NS", ylab = "Mean dN/dS",
                                y.text.col = TRUE, rotate = TRUE, palette = snp_nsgenes_sub_col,
                                dot.size = 3, sorting = "descending", ggtheme = theme_pubr()) +
  scale_y_continuous(limits = c(0, 1)) +
  geom_hline(yintercept = mean(snp_nsgenes$wmis_cv), linetype = 2, linewidth = 0.4) +
  geom_hline(yintercept = 1, linetype = 1, linewidth = 0.4) +
  font("legend.title", size = 14) +
  theme(legend.text = element_text(size = 14),
        axis.text = element_text(size = 16),
        axis.title = element_text(size = 20))
ggsave("article_profiling/5-SNPs/dndscv/results/plot/plot_dnds_nsgenesprod.pdf", snp_plot_nsgensub, height = 11, width = 13)

@@

# Boxplot global dN/dS ratios by phylum
snp_plot_estim_phy <- ggboxplot(snp_estim_phy, x = "phylum", y = "dnds_glob", color = "phylum",
                                order = snp_estim_sort_phy$phylum, palette = snp_estim_sort_phy$color,
                                add = "jitter", size = 0.75, legend = "none",
                                add.params = list(size = 3, alpha = 0.75)) +
  geom_hline(yintercept = mean(snp_estim_phy$dnds_glob), linetype = 2, linewidth = 0.4) +
  geom_hline(yintercept = 1, linetype = 1, linewidth = 0.4) +
  labs(x = "Phylum", y = "Global normalized dN/dS ratio") +
  stat_compare_means(method = "anova", label.y = 1.6, size = 6) +
  stat_compare_means(comparisons = list(c("Proteobacteria", "Actinobacteriota"),
                                        c("Proteobacteria", "Acidobacteriota"),
                                        c("Actinobacteriota", "Acidobacteriota")),
                     label = "p.format", method = "t.test", size = 5) +
  theme(axis.text = element_text(size = 16),
        axis.title = element_text(size = 20))
ggsave("article_profiling/5-SNPs/dndscv/results/plot/plot_dnds_estim_phy.pdf", snp_plot_estim_phy, height = 10, width = 7)

@@

# Barplot global dN/dS ratios by MAG bins
snp_plot_estim_bin <- ggbarplot(snp_estim_bin, x = "bin", y = "dnds_glob", color = "phylum",
                                order = snp_estim_sort_bin$bin, palette = tree_colors,
                                fill = "phylum", rotate = TRUE) +
  geom_hline(yintercept = 1, linetype = 1, linewidth = 0.4) +
  labs(x = "MAG bin", y = "Global normalized dN/dS ratio") +
  font("legend.title", size = 14) +
  theme(legend.text = element_text(size = 14),
        axis.text = element_text(size = 16),
        axis.title = element_text(size = 20))
ggsave("article_profiling/5-SNPs/dndscv/results/plot/plot_dnds_estim_bin.pdf", snp_plot_estim_bin, height = 11, width = 8)

@@

```{r fig4}
# Merge plots using patchwork
fig_4 <- wrap_plots(snp_plot_estim_bin, snp_plot_estim_phy, design = "AABB", heights = c(1, 0.9)) +
  plot_annotation(tag_levels = list(c("a", "b"))) &
  theme(plot.tag = element_text(size = 22, face = "bold"))
ggsave("article_profiling/fig_4.pdf", fig_4, height = 12, width = 15)
ggsave("article_profiling/fig_4.png", fig_4, height = 12, width = 15)

fig_5a <- wrap_plots(snp_plot_nsgenes, snp_plot_nsgensub, design = "AABB", heights = c(0.9, 1)) +
  plot_annotation(tag_levels = list(c("a", "b"))) &
  theme(plot.tag = element_text(size = 22, face = "bold"))
ggsave("article_profiling/fig_5a.pdf", fig_5a, height = 11, width = 19)
ggsave("article_profiling/fig_5a.png", fig_5a, height = 11, width = 19)

fig_5b <- wrap_plots(snp_plot_psgenes, snp_plot_kos, design = "AABBB", heights = c(0.9, 1)) +
  plot_annotation(tag_levels = list(c("c", "d"))) &
  theme(plot.tag = element_text(size = 22, face = "bold"))
ggsave("article_profiling/fig_5b.pdf", fig_5b, height = 12, width = 23)
ggsave("article_profiling/fig_5b.png", fig_5b, height = 12, width = 23)

#sup 3
fig_s3 <- wrap_plots(snp_plot_psbins, widths = c(1)) +
  plot_annotation(tag_levels = list(c("a", "", ""))) &
  theme(plot.tag = element_text(size = 22, face = "bold"))
#ggsave("article_profiling/fig_s3.pdf", fig_s3, height = 11, width = 17)
#ggsave("article_profiling/fig_s3.png", fig_s3, height = 11, width = 17)
```

>>>>

suppressMessages(library(ape)) #tree basics
suppressMessages(library(treeio)) #join trees
suppressMessages(library(ggtree)) #tree plotting
suppressMessages(library(ggplot2)) #plots
suppressMessages(library(ggtreeExtra)) #geom_fruit
suppressMessages(library(ggnewscale)) #scales in ggplot2
suppressMessages(library(ggstar)) #geom_star for ggplot2
suppressMessages(library(tidyverse)) #strings, dataframes processing
suppressMessages(library(magrittr)) #pipes %>%
suppressMessages(library(tidytree)) #trees as dataframes
suppressMessages(library(RColorBrewer)) #color palettes
suppressWarnings(library(colorspace)) #colors manipulation
suppressWarnings(library(patchwork)) #side-by-side plotting
suppressWarnings(library(openxlsx)) #read excel
suppressWarnings(library(ComplexHeatmap)) #heatmap
suppressWarnings(library(PCAtools)) #pca
suppressWarnings(library(ggbiplot)) #pca plots
suppressWarnings(library(agricolae)) #fisher's test
suppressWarnings(library(ggpubr)) #stats & plots
suppressWarnings(library(dndscv)) #dN/dS of snps
suppressWarnings(library(clusterProfiler)) #gene enrichment
suppressWarnings(library(circlize)) #circular viz
suppressWarnings(library(ggbeeswarm)) #scatter plots
```

## Prepare and merge MAG-SMAG

@@

### 1.1: In-depth biomining MAGs vs SMAG

@@

### 1.2: MAGs QC

@@

### 1.3: Phylogenetic tree

@@

## Section 2: Abundances

```{r abund}
# Read matrix of raw abundances
abu_depths <- read.csv("article_profiling/3-abundances/raw/bin_refined_depths_summary.tsv", sep = "\t")

# Filter list to S15 sample
abu_depths <- abu_depths %>%
  dplyr::filter(!grepl("Unbinned", bin)) %>%
  dplyr::filter(grepl("S15", bin)) %>%
  dplyr::mutate(bin = str_replace_all(bin, ".fa", "")) %>%
  dplyr::select(bin, S15)

# Add phylum
abu_depths <- left_join(abu_depths, dplyr::select(gtdb_taxonomy, label, phylum),
                        by = join_by(bin == label))

# Dotplot of global abundances by MAG size
abu_plot_depth <- abu_depths %>%
  dplyr::mutate(bin = str_replace_all(bin, "SPAdes-MaxBin2Refined-", "")) %>% #trim bin names
  dplyr::mutate(bin = str_replace_all(bin, "SPAdes-MetaBAT2Refined-", "")) %>%
  ggplot(aes(x = S15, y = reorder(bin, S15), color = phylum)) +
  geom_point(aes(size = S15), alpha = 0.7) +
  scale_size(range = c(0, 10)) +
  scale_color_manual(values = tree_colors) +
  scale_x_log10(expand = c(0.1, 0)) + #add x-axis space
  scale_y_discrete(expand = c(0.025, 0)) + #add y-axis space
  guides(color = guide_legend(title = "Phylum", ncol = 1), size = "none") +
  labs(x = "Abundance (mean coverage)", y = "MAG") +
  theme_classic() +
  theme(legend.text = element_text(size = 11),
        legend.title = element_text(size = 13),
        axis.text = element_text(size = 12),
        axis.title = element_text(size = 14))
ggsave("article_profiling/3-abundances/plots/plot_abu_rank.pdf", abu_plot_depth, height = 8, width = 6)

# Define top phyla
abu_top_phy <- c("Acidobacteriota", "Actinobacteriota", "Proteobacteria")

# Boxplot of abundances by top phyla
abu_plot_depth.2 <- abu_depths %>%
  dplyr::filter(phylum %in% abu_top_phy) %>%
  dplyr::mutate(phylum = factor(phylum, levels = c("Actinobacteriota", "Acidobacteriota", "Proteobacteria"))) %>% #sort
  ggplot(aes(x = phylum, y = S15, color = phylum)) +
  geom_quasirandom(aes(size = S15), alpha = 0.7) +
  scale_size(range = c(0, 10)) +
  scale_y_log10() +
  geom_boxplot(fill = NA, outlier.size = 0, outliers = FALSE) +
  scale_color_manual(values = c("#98d277", "#a6cee3", "#b15928")) +
  guides(size = "none", color = "none") +
  stat_compare_means(method = "anova", label.y = 3.8, size = 4.5) + #pos by log10
  stat_compare_means(method = "t.test", label = "p.format", ref.group = ".all.", vjust = 0.25,
                     comparisons = list(c("Actinobacteriota", "Acidobacteriota"),
                                        c("Actinobacteriota", "Proteobacteria"),
                                        c("Acidobacteriota", "Proteobacteria")),
                     label.y = c(3.3, 3.45, 3.64), bracket.size = 0.25, size = 4) +
  labs(x = "Phylum", y = "Abundance (mean coverage)") +
  theme_classic() +
  theme(axis.text = element_text(size = 12),
        axis.title = element_text(size = 14))
ggsave("article_profiling/3-abundances/plots/plot_abu_phy.pdf", abu_plot_depth.2, height = 7, width = 5)
```

@@

```{r fig1}
# Merge plots using patchwork
fig_1a <- wrap_plots(plot_mags_qc, plot_depth_sgbs_biom, plot_tree_s15_md,
                     design = "ACCC\nBCCC", widths = c(0.6, 1)) +
  plot_annotation(tag_levels = list(c("a", "b", "c"))) &
  theme(plot.tag = element_text(size = 16, face = "bold"))
ggsave("article_profiling/fig_1a.pdf", fig_1, height = 8.5, width = 13)
#ggsave("article_profiling/fig_1a.png", fig_1, height = 8.5, width = 13)
fig_1b <- wrap_plots(abu_plot_depth, abu_plot_depth.2, widths = c(0.65, 1)) +
  plot_annotation(tag_levels = list(c("d", "e"))) &
  theme(plot.tag = element_text(size = 16, face = "bold"))
ggsave("article_profiling/fig_1b.pdf", fig_1b, height = 8, width = 11)
#ggsave("article_profiling/fig_1b.png", fig_1b, height = 8, width = 11)
#compose
fig_1 <- wrap_plots(fig_1a / fig_1b) +
  plot_annotation(tag_levels = list(c("a", "b", "c", "d", "e"))) &
  theme(plot.tag = element_text(size = 16, face = "bold"))
ggsave("article_profiling/fig_1.pdf", fig_1, height = 16, width = 13)
ggsave("article_profiling/fig_1.png", fig_1, height = 16, width = 13)
rm(fig_1a, fig_1b) #clean memory

fig_2 <- wrap_plots(plot_depth_sgbs_both, plot_depth_spec_eco,
                    plot_depth_kggbs_biom, design = "AAABBCC") +
  plot_annotation(tag_levels = list(c("a", "", "b", "c"))) &
  theme(plot.tag = element_text(size = 18, face = "bold"))
ggsave("article_profiling/fig_2.pdf", fig_2, height = 8, width = 16)
ggsave("article_profiling/fig_2.png", fig_2, height = 8, width = 16)

@@

## Section 3: Functional Annotation

### 3.1: Process KEGG KOs

@@

### 3.2: Heatmap

@@

### 3.3: PCA (**discarded**)

@@

### 3.4: Bubble/boxplots

@@

## Section 4: Evolutionary analysis

@@

# Violin plot for NS genes with global mutations
snp_plot_nsgenes <- snp_nsgenes %>%
  ggviolin(x = "pathway", y = "wmis_cv", color = "pathway", add = "jitter", shape = "pathway",
           order = snp_sort_ns$pathway, palette = snp_sort_ns$color, legend = "none",
           size = 0.55, add.params = list(size = 3, alpha = 0.7)) +
  scale_y_continuous(limits = c(-0.2, 1.5)) +
  geom_hline(yintercept = 1, linetype = 1, linewidth = 0.4) +
  geom_hline(yintercept = mean(snp_nsgenes$wmis_cv), linetype = 2, linewidth = 0.4) +
  labs(x = "Pathway", y = "dN/dS") +
  stat_compare_means(method = "anova", label.y = 1.475, size = 6) +
  stat_compare_means(method = "t.test", label = "p.format", vjust = 0.25,
                     comparisons = list(c("Sulfur", "Copper"), c("Sulfur", "Iron"), c("Copper", "Iron")),
                     label.y = c(1.2, 1.27, 1.36), bracket.size = 0.25, size = 5) +
  theme(axis.text = element_text(size = 16),
        axis.title = element_text(size = 18))
ggsave("article_profiling/5-SNPs/dndscv/results/plot/plot_dnds_nsgenes.pdf", snp_plot_nsgenes, height = 9, width = 6)

@@

# Dotplot of grouped products of NS genes
snp_plot_nsgensub <- snp_nsgenes_sub %>%
  dplyr::rename(Pathway = pathway) %>% #rename col.
  ggdotchart(x = "product", y = "dnds", color = "Pathway", y.text.col = TRUE, rotate = TRUE,
             palette = snp_nsgenes_sub_col, dot.size = 3, sorting = "descending") + #add = "segments"
  scale_y_continuous(limits = c(0, 1)) +
  geom_hline(yintercept = mean(snp_nsgenes$wmis_cv), linetype = 2, linewidth = 0.4) +
  labs(x = "NS gene products", y = "Mean dN/dS") +
  font("legend.title", size = 15) +
  theme(legend.text = element_text(size = 14),
        axis.text = element_text(size = 15),
        axis.title = element_text(size = 18))
ggsave("article_profiling/5-SNPs/dndscv/results/plot/plot_dnds_nsgenesprod.pdf", snp_plot_nsgensub, height = 10, width = 11)

@@

# Boxplot global dN/dS ratios by phylum
snp_plot_estim_phy <- snp_estim_phy %>%
  ggboxplot(x = "phylum", y = "dnds_glob", color = "phylum", add = "jitter",
            order = snp_estim_sort_phy$phylum, palette = snp_estim_sort_phy$color,
             size = 0.55, legend = "none", add.params = list(size = 3.25, alpha = 0.7)) +
  geom_hline(yintercept = mean(snp_estim_phy$dnds_glob), linetype = 2, linewidth = 0.4) +
  geom_hline(yintercept = 1, linetype = 1, linewidth = 0.4) +
  labs(x = "Phylum", y = "Global normalized dN/dS ratio") +
  stat_compare_means(method = "anova", label.y = 1.475, size = 6) +
  stat_compare_means(method = "t.test", label = "p.format", vjust = 0.25,
                     comparisons = list(c("Proteobacteria", "Actinobacteriota"),
                                        c("Proteobacteria", "Acidobacteriota"),
                                        c("Actinobacteriota", "Acidobacteriota")),
                     label.y = c(1.25, 1.325, 1.4), bracket.size = 0.25, size = 5) +
  theme(axis.text = element_text(size = 15),
        axis.title = element_text(size = 18))
ggsave("article_profiling/5-SNPs/dndscv/results/plot/plot_dnds_estim_phy.pdf", snp_plot_estim_phy, height = 9, width = 6)

@@

# Barplot global dN/dS ratios by MAG bins
snp_plot_estim_bin <- snp_estim_bin %>%
  dplyr::rename(Phylum = phylum) %>% #rename col.
  ggbarplot(x = "bin", y = "dnds_glob", color = "Phylum", fill = "Phylum", legend = "right",
            order = snp_estim_sort_bin$bin, palette = tree_colors, rotate = TRUE) +
  geom_hline(yintercept = 1, linetype = 1, linewidth = 0.4) +
  labs(x = "MAG", y = "Global normalized dN/dS ratio") +
  font("legend.title", size = 15) +
  theme(legend.text = element_text(size = 14),
        axis.text = element_text(size = 15),
        axis.title = element_text(size = 18))
ggsave("article_profiling/5-SNPs/dndscv/results/plot/plot_dnds_estim_bin.pdf", snp_plot_estim_bin, height = 9, width = 7)

@@

```{r fig4}
# Merge plots using patchwork
fig_4 <- wrap_plots(snp_plot_estim_bin, snp_plot_estim_phy, widths = c(0.6, 1)) +
  plot_annotation(tag_levels = list(c("a", "b"))) &
  theme(plot.tag = element_text(size = 20, face = "bold"))
ggsave("article_profiling/fig_4.pdf", fig_4, height = 10, width = 13)
ggsave("article_profiling/fig_4.png", fig_4, height = 10, width = 13)

fig_5a <- wrap_plots(snp_plot_nsgenes, snp_plot_nsgensub, widths = c(1, 1)) +
  plot_annotation(tag_levels = list(c("a", "b"))) &
  theme(plot.tag = element_text(size = 20, face = "bold"))
ggsave("article_profiling/fig_5a.pdf", fig_5a, height = 10, width = 15)
#ggsave("article_profiling/fig_5a.png", fig_5a, height = 10, width = 15)
fig_5b <- wrap_plots(snp_plot_psgenes, snp_plot_kos, design = "AABBB", heights = c(0.9, 1)) +
  plot_annotation(tag_levels = list(c("c", "d"))) &
  theme(plot.tag = element_text(size = 22, face = "bold"))
ggsave("article_profiling/fig_5b.pdf", fig_5b, height = 12, width = 22)
#ggsave("article_profiling/fig_5b.png", fig_5b, height = 12, width = 22)
#compose
fig_5 <- wrap_plots(wrap_plots(fig_5a, plot_spacer(), widths = c(1, 0.35)) / fig_5b) +
  plot_annotation(tag_levels = list(c("a", "b", "c", "d"))) &
  theme(plot.tag = element_text(size = 22, face = "bold"))
ggsave("article_profiling/fig_5.pdf", fig_5, height = 22, width = 22)
ggsave("article_profiling/fig_5.png", fig_5, height = 22, width = 22)
rm(fig_5a, fig_5b) #clean memory

#sup 3
#fig_s3 <- wrap_plots(snp_plot_psbins, widths = c(1)) +
#  plot_annotation(tag_levels = list(c("a", "", ""))) &
#  theme(plot.tag = element_text(size = 22, face = "bold"))
#ggsave("article_profiling/fig_s3.pdf", fig_s3, height = 11, width = 17)
#ggsave("article_profiling/fig_s3.png", fig_s3, height = 11, width = 17)
```
